Harnessing AI to annotate the severity of all phenotypic abnormalities within the Human Phenotype Ontology

### 2024-05-14 docs/_tex/manuscript.tex
Update manuscript to address issues: #1 #2 #3 #4 #5 #6 #7 #8 #9

To date, the HPO has largely relied on manual curation by experts
including clinicians, clinical geneticists and researchers. Although
this approach ensures the quality and accuracy of the ontology, it is
time-consuming and labour-intensive. As a result, less than 1\% of terms
within the HPO have metadata pertaining to their features such as time
course and severity. Artificial intelligence (AI) capabilities have
advanced considerably in recent years, presenting new opportunities to
integrate natural language processing technologies into assisting in the
curation process. Specifically, there have recently been considerable
advances in large language model (LLM) and their application to
biomedical problems, in some cases performing as well or better than
human clinicians on standardised medical exams and patient diagnosis
tasks (Bolton et al., n.d.; Cheng et al., 2023; Gu et al., 2021; Labrak
et al., n.d.; Luo et al., 2022; McDuff et al., n.d.; O'Neil et al.,
n.d.; Shin et al., 2020; Singhal et al., n.d., 2023, 2023; Van Veen et
al., 2024; Zhang et al., n.d.). Recent work has demonstrated that the
Generative Pre-trained Transformer 4 (GPT-4) foundation model (OpenAI et
al., n.d.), when combined with strategic prompt engineering, can
outperform even specialist LLMs that are explicitly fine-tuned for
biomedical tasks (Nori et al., n.d.). In a landmark achievement, GPT-4
was the first LLM to surpass a score of 90\% in the United States
Medical Licensing Examination (USML) (Nori et al., n.d.).

@@

To assess annotation consistency, 793 phenotypes underwent duplicate
annotations. We employed two different metrics to determine the
\emph{consistency rate}. The first, less stringent metric, defined
consistency as the duplicate annotations being either `always' and
`often', or `never' and `rarely'. The second, more stringent metric,
required exact agreement in annotation occurrences, e.g.~`always' and
`always'. For the less stringent metric, duplicated phenotypes were
annotated consistently at a rate of at least 80\%, and for the more
stringent metric, the lowest consistency rate was 57\% for congenital
onset. An example of where annotations were inconsistent was for the HPO
term `Acute leukaemia'. One time, GPT-4 annotated it as often causing
impaired mobility, giving the justification that `weakness and fatigue
from leukaemia and its treatment can impair mobility'. The other time,
GPT-4 annotated it as rarely causing impaired mobility, giving the
justification that `acute leukaemia rarely impairs mobility directly'.
Despite specifying in the prompt for GPT-4 not to take into
consideration indirect effects, here is an example of where it failed to
do so.

@@

@@

The creation and annotation of biomedical knowledge has traditionally
relied on manual or semi-manual curation by human experts (Gargano et
al., 2024; Köhler et al., 2021; Mungall et al., 2017; Ochoa et al.,
2021; Putman et al., 2024). Performing such manual curation and review
tasks at scale is often infeasible for human biomedical experts given
limited time and resources. LLMs have the capacity to effectively
encode, retrieve, and synthesise vast amounts of diverse information in
a highly scalable manner (OpenAI et al., n.d.; Singhal et al., 2023; Van
Veen et al., 2024). This makes them powerful tools that can be applied
in a rapidly expanding variety of scenarios, including medical practice,
research and data curation (Caufield et al., n.d.; O'Neil et al., n.d.;
Pan et al., n.d.; Singhal et al., 2023; Toro et al., n.d.).

Here, we introduce a novel framework to leverage the current
best-in-class LLM, GPT-4 (OpenAI et al., n.d.), to systematically
annotate the severity of 17502 phenotypic abnormalities within the HPO.
By employing advanced AI capabilities, we have demonstrated the
feasibility of automating this process, significantly enhancing
efficiency without substantially compromising accuracy. Our validation
approach yielded a high true positive rate exceeding 88\% across the
phenotypes tested. Furthermore, our approach can be readily adapted and
scaled to accommodate the growing volume of phenotypic data.

@@

In addition to rapidly synthesising and summarising vast amounts of
information, LLMs can also be steered to provide justifications for each
particular response. This makes LLMs amenable to direct interrogation as
a means of recovering explainability, especially when designed to retain
information about previous requests and interactions as they use these
to iteratively improve and update their predictions (Janik, n.d.). This
represents a categorical advance over traditional natural language
processing models based on more shallow forms of statistical or machine
learning (e.g.~Term Frequency-Inverse Document Frequency (Jones, 1972),
Word2vec (Mikolov et al., n.d.)) which lack the ability to provide
chains of causal reasoning to justify their predictions. This highlights
the fundamental trade-off between simpler models with high
explainability (the ability humans to understand the inner workings of
the model) but low interpretability (the ability of humans to trace the
decision process of the model, analogous to human `reasoning'), and
deeper more complex models with low explainability but high
interpretability (Marcinkevičs \& Vogt, n.d.).

A key contribution of our study is the introduction of a quantitative
severity scoring system that integrates both the nature of the phenotype
outcome and the frequency of its occurrence. By encoding the concept of
severity in this way, we are able to prioritise phenotypes based on
their impact on patients. The methodology allowed us to transition from
low-throughput qualitative assessments of severity (e.g. Lazarin et al.
(2014)) to high-throughput quantitative assessments of severity. One of
the most severe phenotypes in the HPO is `Fetal akinesia sequence' (FAS;
HP:0001989, severity score= 50.7), and extremely rare condition that is
almost always lethal. FAS is a complex, multi-system phenotype that can
be caused by at least 24 different genetic disorders. Despite the
complex and heterogenous aetiology of this phenotype, GPT-4 was able to
provide accurate annotations alongside explainable justifications for
those annotations (see Table~\ref{tbl-fas}). For example, this phenotype
almost always results in death, either \emph{in utero} or shortly after
birth. Not only did GPT-4 correctly provide the annotation death as
`always', when asked whether FAS causes sensory impairments it provided
the response `always' with the justification `Fetal akinesia sequence
typically results in severe sensory impairment due to neurodevelopmental
disruption.' Neurodevelopmental disruption is indeed a hallmark
component of FAS (e.g.~hydrocephalus, cerebellar hypoplasia) that causes
severe impairments across multiple sensory systems (Chen, 2012). This
demonstrates that GPT-4 was able to recover the correct chain of
causality from phenotype to clinical outcome.

Our findings highlight the potential of this next generation of natural
language processing technologies in significantly contributing to the
automation and refinement of data curation in biomedical research. These
results have a large number of useful real-world applications, such as
prioritising gene therapy candidates (Murphy et al., n.d.) and guiding
clinical decision-making in rare diseases. As the HPO continues to grow
over time, this approach can easily be repeated to keep pace with the
rapidly evolving biomedical landscape. Furthermore, this approach can be
extended to produce different sets of annotations or be used with any
other ontology. Additional use cases include gathering data on the
prevalence of each phenotype to approximate their social and financial
costs.

@@

We wrote a Python script to iteratively query GPT-4 via the OpenAI
application programming interface (API). The ultimately yielded
consistently formatted annotations for 17502 terms within the HPO. Our
annotation framework was developed based on previously defined criteria
for classifying disease severity (Lazarin et al., 2014). We sought to
evaluate whether each phenotype directly caused a given severity-related
clinical characteristic, including: intellectual disability, death,
impaired mobility, physical malformations, blindness, sensory
impairments, immunodeficiency, cancer, reduced fertility, and/or had a
congenital onset. Through prompt engineering we found that the
performance of GPT-4 improved when we incorporated a scale associated
with each effect and required a justification for each response. We
asked how frequently the given phenotype directly causes each clinical
characteristic - whether it never, rarely, often, or always occurred.
Furthermore, our prompt design revealed that the optimal trade-off
between the number of phenotypes and performance (in terms of producing
the desired annotations, and adhering to the formatting requirements)
was achieved when inputting no more than two or three phenotypes per
prompt. An example prompt can be seen in Figure~\ref{fig-occurrence}.
Only 2 phenotypes were included per prompt in order to 1) avoid
exceeding per-query token limits, and 2) prevent the breakdown of GPT-4
performance due to long-form text input, which is presently a known
limitation common to many LLMs including GPT-4 (Wei et al., n.d.).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1194}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.7711}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1095}}@{}}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Clinical characteristic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
HPO queries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
True positive HPO IDs
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Intellectual disability & Intellectual disability; Mental deterioration
& 19 \\
Impaired mobility & Gait disturbance; Diminished movement; mobility &
319 \\
Physical malformations & malformation & 78 \\
Blindness & blindness & 1 \\
Sensory impairments & Abnormality of vision; Abnormality of the sense of
smell; Abnormality of taste sensation; Somatic sensory dysfunction;
Hearing abnormality & 252 \\
Immunodeficiency & Immunodeficiency; Impaired antigen-specific response
& 29 \\
Cancer & Cancer; malignant; carcinoma & 56 \\
Reduced fertility & Decreased fertility; Hypogonadism & 9 \\

@@

\begin{itemize}
\item
  \(p\) : a phenotype in the HPO.
\item
  \(j\) : the identity of a given annotation metric (i.e.~clinical
  characteristic, such as `intellectual disability' or `congenital
  onset').
\item
  \(W_j\): the assigned weight of metric \(j\).
\item
  \(F_j\): the maximum possible value for metric \(j\) (equivalent
  across all \(j\)).
\item
  \(F_{pj}\) : the numerically encoded value of annotation metric \(j\)
  for phenotype \(p\).
\item
  \(NSS_p\): The final composite severity score for phenotype \(p\)
  after applying normalisation to align values to a 0-100 scale and
  ensure equivalent meaning regardless of which other phenotypes are
  being analysed in addition to \(p\). This allows for direct
  comparability of severity scores across studies with different sets of
  phenotypes.
\end{itemize}

@@

@@

\bibitem[\citeproctext]{ref-boltonBioMedLM7BParameter2024}
Bolton, E., Venigalla, A., Yasunaga, M., Hall, D., Xiong, B., Lee, T.,
et al. (n.d.). BioMedLM: A 2.7B parameter language model trained on
biomedical text. \url{https://doi.org/10.48550/arXiv.2403.18421}

\bibitem[\citeproctext]{ref-caufieldStructuredPromptInterrogation2023}
Caufield, J. H., Hegde, H., Emonet, V., Harris, N. L., Joachimiak, M.
P., Matentzoglu, N., et al. (n.d.). Structured prompt interrogation and
recursive extraction of semantics (SPIRES): A method for populating
knowledge bases using zero-shot learning.
\url{https://doi.org/10.48550/arXiv.2304.02711}

@@

\bibitem[\citeproctext]{ref-janikAspectsHumanMemory2024}
Janik, R. A. (n.d.). Aspects of human memory and large language models.
\url{https://doi.org/10.48550/arXiv.2311.03839}

@@

\bibitem[\citeproctext]{ref-labrakBioMistralCollectionOpenSource2024}
Labrak, Y., Bazoge, A., Morin, E., Gourraud, P.-A., Rouvier, M., \&
Dufour, R. (n.d.). BioMistral: A collection of open-source pretrained
large language models for medical domains.
\url{https://doi.org/10.48550/arXiv.2402.10373}

@@

\bibitem[\citeproctext]{ref-marcinkevicsInterpretabilityExplainabilityMachine2023}
Marcinkevičs, R., \& Vogt, J. E. (n.d.). Interpretability and
explainability: A machine learning zoo mini-tour.
\url{https://doi.org/10.48550/arXiv.2012.01805}

\bibitem[\citeproctext]{ref-mcduffAccurateDifferentialDiagnosis2023}
McDuff, D., Schaekermann, M., Tu, T., Palepu, A., Wang, A., Garrison,
J., et al. (n.d.). Towards accurate differential diagnosis with large
language models. \url{https://doi.org/10.48550/arXiv.2312.00164}

\bibitem[\citeproctext]{ref-mikolovEfficientEstimationWord2013}
Mikolov, T., Chen, K., Corrado, G., \& Dean, J. (n.d.). Efficient
estimation of word representations in vector space.
\url{https://doi.org/10.48550/arXiv.1301.3781}

@@

\bibitem[\citeproctext]{ref-murphyIdentificationCellTypespecific2023}
Murphy, K. B., Gordon-Smith, R., Chapman, J., Otani, M., Schilder, B.
M., \& Skene, N. G. (n.d.). Identification of cell type-specific gene
targets underlying thousands of rare diseases and subtraits.
\url{https://doi.org/10.1101/2023.02.13.23285820}

\bibitem[\citeproctext]{ref-noriCanGeneralistFoundation2023}
Nori, H., Lee, Y. T., Zhang, S., Carignan, D., Edgar, R., Fusi, N., et
al. (n.d.). Can generalist foundation models outcompete special-purpose
tuning? Case study in medicine.
\url{https://doi.org/10.48550/arXiv.2311.16452}

\bibitem[\citeproctext]{ref-oneilPhenomicsAssistantInterface2024}
O'Neil, S. T., Schaper, K., Elsarboukh, G., Reese, J. T., Moxon, S. A.
T., Harris, N. L., et al. (n.d.). Phenomics assistant: An interface for
LLM-based biomedical knowledge graph exploration.
\url{https://doi.org/10.1101/2024.01.31.578275}

@@

\bibitem[\citeproctext]{ref-openaiGPT4TechnicalReport2024}
OpenAI, Achiam, J., Adler, S., Agarwal, S., Ahmad, L., Akkaya, I., et
al. (n.d.). GPT-4 technical report.
\url{https://doi.org/10.48550/arXiv.2303.08774}

\bibitem[\citeproctext]{ref-panLargeLanguageModels2023}
Pan, J. Z., Razniewski, S., Kalo, J.-C., Singhania, S., Chen, J.,
Dietze, S., et al. (n.d.). Large language models and knowledge graphs:
Opportunities and challenges.
\url{https://doi.org/10.48550/arXiv.2308.06374}

@@

\bibitem[\citeproctext]{ref-singhalExpertLevelMedicalQuestion2023}
Singhal, K., Tu, T., Gottweis, J., Sayres, R., Wulczyn, E., Hou, L., et
al. (n.d.). Towards expert-level medical question answering with large
language models. \url{https://doi.org/10.48550/arXiv.2305.09617}

@@

\bibitem[\citeproctext]{ref-toroDynamicRetrievalAugmented2023}
Toro, S., Anagnostopoulos, A. V., Bello, S., Blumberg, K., Cameron, R.,
Carmody, L., et al. (n.d.). Dynamic retrieval augmented generation of
ontologies using artificial intelligence (DRAGON-AI).
\url{https://doi.org/10.48550/arXiv.2312.10904}

@@

\bibitem[\citeproctext]{ref-weiLongformFactualityLarge2024}
Wei, J., Yang, C., Song, X., Lu, Y., Hu, N., Huang, J., et al. (n.d.).
Long-form factuality in large language models. Retrieved from
\url{http://arxiv.org/abs/2403.18802}

\bibitem[\citeproctext]{ref-zhangBiomedGPTUnifiedGeneralist2023}
Zhang, K., Yu, J., Yan, Z., Liu, Y., Adhikarla, E., Fu, S., et al.
(n.d.). BiomedGPT: A unified and generalist biomedical generative
pre-trained transformer for vision, language, and multimodal tasks.
\url{https://doi.org/10.48550/arXiv.2305.17100}

@@

@@

>>>>

To date, the HPO has largely relied on manual curation by experts
including clinicians, clinical geneticists and researchers. Although
this approach ensures the quality and accuracy of the ontology, it is
time-consuming and labour-intensive. As a result, less than 1\% of terms
within the HPO have metadata pertaining to their features such as time
course and severity. Artificial intelligence (AI) capabilities have
advanced considerably in recent years, presenting new opportunities to
integrate natural language processing technologies into assisting in the
curation process. Specifically, there have recently been considerable
advances in large language model (LLM) and their application to
biomedical problems, in some cases performing as well or better than
human clinicians on standardised medical exams and patient diagnosis
tasks (Bolton et al., 2024; Cheng et al., 2023; Gu et al., 2021; Labrak
et al., 2024; Luo et al., 2022; McDuff et al., 2023; O'Neil et al.,
2024; Shin et al., 2020; Singhal, Azizi, et al., 2023, 2023; Singhal,
Tu, et al., 2023; Van Veen et al., 2024; Zhang et al., 2023). Recent
work has demonstrated that the Generative Pre-trained Transformer 4
(GPT-4) foundation model (OpenAI et al., 2024), when combined with
strategic prompt engineering, can outperform even specialist LLMs that
are explicitly fine-tuned for biomedical tasks (Nori et al., 2023). In a
landmark achievement, GPT-4 was the first LLM to surpass a score of 90\%
in the United States Medical Licensing Examination (USML) (Nori et al.,
2023).

@@

To assess annotation consistency, 793 we queried GPT-4 with a subset of
the HPO phenotypes multiple times. We employed two different metrics to
determine the \emph{consistency rate}. The first, less stringent metric,
defined consistency as the duplicate annotations being either `always'
and `often', or `never' and `rarely'. The second, more stringent metric,
required exact agreement in annotation occurrences, e.g.~`always' and
`always'. For the less stringent metric, duplicated phenotypes were
annotated consistently at a rate of at least 80\%, and for the more
stringent metric, the lowest consistency rate was 57\% for congenital
onset. An example of where annotations were inconsistent was for the HPO
term `Acute leukaemia'. One time, GPT-4 annotated it as often causing
impaired mobility, giving the justification that `weakness and fatigue
from leukaemia and its treatment can impair mobility'. The other time,
GPT-4 annotated it as rarely causing impaired mobility, giving the
justification that `acute leukaemia rarely impairs mobility directly'.
Despite specifying in the prompt for GPT-4 not to take into
consideration indirect effects, this is an example of where it failed to
do so.

We also reasoned that GPT-4 would be better able to give consistent
answers for more specific phenotypes lower in the ontology, as they are
more likely to have a single cause. We found that the stringent
consistency rate did indeed significantly improve with greater HPO
ontology depth (\(X_{Pearson}^2\)=22.17, \(\hat{V}_{Cramer}\)=0.03,
p=0.05). See Figure~\ref{fig-consist-vs-ontLvl} for a visual
representation of this relationship.

@@

\subsubsection{Severity classes}\label{severity-classes}

While the continuous severity score is a helpful metric, there may be
some use cases where a categorical classification of severity is more
immediately useful. In work by Lazarin et al. (2014), the authors
defined severity classed using a simple decision tree based on the
individual severity annotations. We approximated this approach using our
GPT-4 annotations. This categorical approach showed a strong degree of
positive correspondence with the continuous severity score
(\(\hat{\omega_{p}^2}\)=0.87, p\textless2.2e-308). In other words,
severity score increased with severity class level (mild \textless{}
moderate \textless{} severe \textless{} profound) as expected. The
distribution of severity classes is shown in
Figure~\ref{fig-severity-class}.

\subsubsection{Correlations between severity
metrics}\label{correlations-between-severity-metrics}

We found that some severity metrics were correlated with one another,
with a mean Pearson correlation of 0.2 across all individual metrics
(see Figure~\ref{fig-metric-corplot}). In particular, blindness and
sensory impairment were highly correlated with one another (r=0.62,
p=0). Some metrics drove the composite severity score more than other,
which is a reflection of both our per-metric weighting scheme, response
type frequencies, and the correlation structure between metrics.
Overall, impaired mobility seemed to be the strongest driver of the
composite severity score with a Pearson correlation of 0.6138757,
followed by intellectual disability (r=0.6) and sensory impairments
(r=0.54).

@@

The creation and annotation of biomedical knowledge has traditionally
relied on manual or semi-manual curation by human experts (Gargano et
al., 2024; Köhler et al., 2021; Mungall et al., 2017; Ochoa et al.,
2021; Putman et al., 2024). Performing such manual curation and review
tasks at scale is often infeasible for human biomedical experts given
limited time and resources. LLMs have the capacity to effectively
encode, retrieve, and synthesise vast amounts of diverse information in
a highly scalable manner (OpenAI et al., 2024; Singhal, Azizi, et al.,
2023; Van Veen et al., 2024). This makes them powerful tools that can be
applied in a rapidly expanding variety of scenarios, including medical
practice, research and data curation (Caufield et al., 2023; O'Neil et
al., 2024; Pan et al., 2023; Singhal, Azizi, et al., 2023; Toro et al.,
2023).

Here, we introduce a novel framework to leverage the current
best-in-class LLM, GPT-4 (OpenAI et al., 2024), to systematically
annotate the severity of 17502 phenotypic abnormalities within the HPO.
By employing advanced AI capabilities, we have demonstrated the
feasibility of automating this process, significantly enhancing
efficiency without substantially compromising accuracy. Our validation
approach yielded a high true positive rate exceeding 88\% across the
phenotypes tested. Furthermore, our approach can be readily adapted and
scaled to accommodate the growing volume of phenotypic data.

@@

In addition to rapidly synthesising and summarising vast amounts of
information, LLMs can also be steered to provide justifications for each
particular response. This makes LLMs amenable to direct interrogation as
a means of recovering explainability, especially when designed to retain
information about previous requests and interactions as they use these
to iteratively improve and update their predictions (Janik, 2024). This
represents a categorical advance over traditional natural language
processing models based on more shallow forms of statistical or machine
learning (e.g.~Term Frequency-Inverse Document Frequency (Jones, 1972),
Word2vec (Mikolov et al., 2013)) which lack the ability to provide
chains of causal reasoning to justify their predictions. This highlights
the fundamental trade-off between simpler models with high
explainability (the ability humans to understand the inner workings of
the model) but low interpretability (the ability of humans to trace the
decision process of the model, analogous to human `reasoning'), and
deeper more complex models with low explainability but high
interpretability (Marcinkevičs \& Vogt, 2023).

A key contribution of our study is the introduction of a quantitative
severity scoring system that integrates both the nature of the phenotype
outcome and the frequency of its occurrence. By encoding the concept of
severity in this way, we are able to prioritise phenotypes based on
their impact on patients. The methodology allowed us to transition from
low-throughput qualitative assessments of severity (e.g. Lazarin et al.
(2014)) to high-throughput quantitative assessments of severity. One of
the most severe phenotypes in the HPO is `Fetal akinesia sequence' (FAS;
HP:0001989, severity score= 50.7), and extremely rare condition that is
almost always lethal. FAS is a complex, multi-system phenotype that can
be caused by at least 24 different genetic disorders. Despite the
complex and heterogeneous aetiology of this phenotype, GPT-4 was able to
provide accurate annotations alongside explainable justifications for
those annotations (see Table~\ref{tbl-fas}). For example, this phenotype
almost always results in death, either \emph{in utero} or shortly after
birth. Not only did GPT-4 correctly provide the annotation death as
`always', when asked whether FAS causes sensory impairments it provided
the response `always' with the justification `Fetal akinesia sequence
typically results in severe sensory impairment due to neurodevelopmental
disruption.' Neurodevelopmental disruption is indeed a hallmark
component of FAS (e.g.~hydrocephalus, cerebellar hypoplasia) that causes
severe impairments across multiple sensory systems (Chen, 2012). This
demonstrates that GPT-4 was able to recover the correct chain of
causality from phenotype to clinical outcome.

Our findings highlight the potential of this next generation of natural
language processing technologies in significantly contributing to the
automation and refinement of data curation in biomedical research. These
results have a large number of useful real-world applications, such as
prioritising gene therapy candidates (Murphy et al., 2023) and guiding
clinical decision-making in rare diseases. As the HPO continues to grow
over time, this approach can easily be repeated to keep pace with the
rapidly evolving biomedical landscape. Furthermore, this approach can be
extended to produce different sets of annotations or be used with any
other ontology. Additional use cases include gathering data on the
prevalence of each phenotype to approximate their social and financial
costs.

@@

We wrote a Python script to iteratively query GPT-4 via the OpenAI
application programming interface (API). The ultimately yielded
consistently formatted annotations for 17502 terms within the HPO. Our
annotation framework was developed based on previously defined criteria
for classifying disease severity (Lazarin et al., 2014). We sought to
evaluate whether each phenotype directly caused a given severity-related
clinical characteristic, including: intellectual disability, death,
impaired mobility, physical malformations, blindness, sensory
impairments, immunodeficiency, cancer, reduced fertility, and/or had a
congenital onset. Through prompt engineering we found that the
performance of GPT-4 improved when we incorporated a scale associated
with each effect and required a justification for each response. We
asked how frequently the given phenotype directly causes each clinical
characteristic - whether it never, rarely, often, or always occurred.
Furthermore, our prompt design revealed that the optimal trade-off
between the number of phenotypes and performance (in terms of producing
the desired annotations, and adhering to the formatting requirements)
was achieved when inputting no more than two or three phenotypes per
prompt. An example prompt can be seen in Figure~\ref{fig-occurrence}.
Only 2 phenotypes were included per prompt in order to 1) avoid
exceeding per-query token limits, and 2) prevent the breakdown of GPT-4
performance due to long-form text input, which is presently a known
limitation common to many LLMs including GPT-4 (Wei et al., 2024).

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4000}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.4000}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2000}}@{}}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Clinical characteristic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
HPO queries
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
True positive HPO IDs
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Intellectual disability & `Intellectual disability'; `Mental
deterioration' & 19 \\
Impaired mobility & `Gait disturbance'; `Diminished movement'; '
mobility' & 319 \\
Physical malformations & `malformation' & 78 \\
Blindness & `blindness' & 1 \\
Sensory impairments & `Abnormality of vision'; `Abnormality of the sense
of smell'; `Abnormality of taste sensation'; `Somatic sensory
dysfunction'; `Hearing abnormality' & 252 \\
Immunodeficiency & `Immunodeficiency'; `Impaired antigen-specific
response' & 29 \\
Cancer & `Cancer'; `malignant'; `carcinoma' & 56 \\
Reduced fertility & `Decreased fertility'; `Hypogonadism' & 9 \\

@@

\begin{itemize}
\item
  \(p\) : a phenotype in the HPO.
\item
  \(j\) : the identity of a given annotation metric (i.e.~clinical
  characteristic, such as `intellectual disability' or `congenital
  onset').
\item
  \(W_j\): the assigned weight of metric \(j\).
\item
  \(F_j\): the maximum possible value for metric \(j\) (equivalent
  across all \(j\)).
\item
  \(F_{pj}\) : the numerically encoded value of annotation metric \(j\)
  for phenotype \(p\).
\item
  \(NSS_p\): the final composite severity score for phenotype \(p\)
  after applying normalisation to align values to a 0-100 scale and
  ensure equivalent meaning regardless of which other phenotypes are
  being analysed in addition to \(p\). This allows for direct
  comparability of severity scores across studies with different sets of
  phenotypes.
\end{itemize}

@@

\subsubsection{Severity classes}\label{severity-classes-1}

The decision tree algorithm used in Lazarin et al. (2014) was adapted
here for use with the GPT-4 annotations. This algorithm first assigned
each annotation metric to a tier, where Tier 1 indicated the most severe
annotation metrics and Tier 4 indicated the least severe annotation
metrics (`death'=1, `intellectual disability'=1, `impaired mobility'=2,
`physical malformations'=2, `blindness'=3, `sensory impairments'=3,
`immunodeficiency'=3, `cancer'=3, `reduced fertility'=4). If a phenotype
often or always caused more than one Tier 1 annotation, it was assigned
a severity class of ``Profound''. If the phenotype often or always
caused only one Tier 1 annotation, it was assigned a severity class of
``Severe''. A ``Severe'' class assignment was also assigned if the
phenotype often or always caused three or more Tier 2 and Tier3
annotations. If the phenotype often or always caused at least one Tier 2
annotation, it was assigned a severity class of ``Moderate''. All
remaining phenotypes were was assigned a severity class of ``Mild''. In
cases where the phenotype mapped to more than one class, only the most
severe class was used. This procedure is implemented within the function
\texttt{HPOExplorer::gpt\_annot\_class}.

\subsubsection{Correlations between severity
metrics}\label{correlations-between-severity-metrics-1}

To assess the correlation structure between each severity metric, as
well as between the composite severity score and each metric, we
computed Pearson correlation coefficients for all pairwise combinations
of these variables using the numerically encoded metric values. The
correlation matrix was visualised using a heatmap, with the colour
intensity representing the strength of the correlation
(Figure~\ref{fig-metric-corplot}).

@@

\bibitem[\citeproctext]{ref-boltonBioMedLM7BParameter2024}
Bolton, E., Venigalla, A., Yasunaga, M., Hall, D., Xiong, B., Lee, T.,
et al. (2024). BioMedLM: A 2.7B parameter language model trained on
biomedical text. \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.2403.18421}

\bibitem[\citeproctext]{ref-caufieldStructuredPromptInterrogation2023}
Caufield, J. H., Hegde, H., Emonet, V., Harris, N. L., Joachimiak, M.
P., Matentzoglu, N., et al. (2023). Structured prompt interrogation and
recursive extraction of semantics (SPIRES): A method for populating
knowledge bases using zero-shot learning. \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.2304.02711}

@@

\bibitem[\citeproctext]{ref-janikAspectsHumanMemory2024}
Janik, R. A. (2024). Aspects of human memory and large language models.
\emph{arXiv}. \url{https://doi.org/10.48550/arXiv.2311.03839}

@@

\bibitem[\citeproctext]{ref-labrakBioMistralCollectionOpenSource2024}
Labrak, Y., Bazoge, A., Morin, E., Gourraud, P.-A., Rouvier, M., \&
Dufour, R. (2024). BioMistral: A collection of open-source pretrained
large language models for medical domains. \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.2402.10373}

@@

\bibitem[\citeproctext]{ref-marcinkevicsInterpretabilityExplainabilityMachine2023}
Marcinkevičs, R., \& Vogt, J. E. (2023). Interpretability and
explainability: A machine learning zoo mini-tour. \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.2012.01805}

\bibitem[\citeproctext]{ref-mcduffAccurateDifferentialDiagnosis2023}
McDuff, D., Schaekermann, M., Tu, T., Palepu, A., Wang, A., Garrison,
J., et al. (2023). Towards accurate differential diagnosis with large
language models. \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.2312.00164}

\bibitem[\citeproctext]{ref-mikolovEfficientEstimationWord2013}
Mikolov, T., Chen, K., Corrado, G., \& Dean, J. (2013). Efficient
estimation of word representations in vector space. \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.1301.3781}

@@

\bibitem[\citeproctext]{ref-murphyIdentificationCellTypespecific2023}
Murphy, K. B., Gordon-Smith, R., Chapman, J., Otani, M., Schilder, B.
M., \& Skene, N. G. (2023). Identification of cell type-specific gene
targets underlying thousands of rare diseases and subtraits.
\emph{medRxiv}. \url{https://doi.org/10.1101/2023.02.13.23285820}

\bibitem[\citeproctext]{ref-noriCanGeneralistFoundation2023}
Nori, H., Lee, Y. T., Zhang, S., Carignan, D., Edgar, R., Fusi, N., et
al. (2023). Can generalist foundation models outcompete special-purpose
tuning? Case study in medicine. \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.2311.16452}

\bibitem[\citeproctext]{ref-oneilPhenomicsAssistantInterface2024}
O'Neil, S. T., Schaper, K., Elsarboukh, G., Reese, J. T., Moxon, S. A.
T., Harris, N. L., et al. (2024). Phenomics assistant: An interface for
LLM-based biomedical knowledge graph exploration. \emph{bioRxiv}.
\url{https://doi.org/10.1101/2024.01.31.578275}

@@

\bibitem[\citeproctext]{ref-openaiGPT4TechnicalReport2024}
OpenAI, Achiam, J., Adler, S., Agarwal, S., Ahmad, L., Akkaya, I., et
al. (2024). GPT-4 technical report. \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.2303.08774}

\bibitem[\citeproctext]{ref-panLargeLanguageModels2023}
Pan, J. Z., Razniewski, S., Kalo, J.-C., Singhania, S., Chen, J.,
Dietze, S., et al. (2023). Large language models and knowledge graphs:
Opportunities and challenges. \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.2308.06374}

@@

@@

\bibitem[\citeproctext]{ref-singhalExpertLevelMedicalQuestion2023}
Singhal, K., Tu, T., Gottweis, J., Sayres, R., Wulczyn, E., Hou, L., et
al. (2023). Towards expert-level medical question answering with large
language models. \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.2305.09617}

\bibitem[\citeproctext]{ref-toroDynamicRetrievalAugmented2023}
Toro, S., Anagnostopoulos, A. V., Bello, S., Blumberg, K., Cameron, R.,
Carmody, L., et al. (2023). Dynamic retrieval augmented generation of
ontologies using artificial intelligence (DRAGON-AI). \emph{arXiv}.
\url{https://doi.org/10.48550/arXiv.2312.10904}

@@

\bibitem[\citeproctext]{ref-weiLongformFactualityLarge2024}
Wei, J., Yang, C., Song, X., Lu, Y., Hu, N., Huang, J., et al. (2024).
Long-form factuality in large language models. \emph{arXiv}. Retrieved
from \url{http://arxiv.org/abs/2403.18802}

\bibitem[\citeproctext]{ref-zhangBiomedGPTUnifiedGeneralist2023}
Zhang, K., Yu, J., Yan, Z., Liu, Y., Adhikarla, E., Fu, S., et al.
(2023). BiomedGPT: A unified and generalist biomedical generative
pre-trained transformer for vision, language, and multimodal tasks.
\emph{arXiv}. \url{https://doi.org/10.48550/arXiv.2305.17100}

@@

\phantomsection\label{cell-fig-consist-vs-ontLvl}
\begin{figure}[H]

\includegraphics{manuscript_files/figure-pdf/fig-consist-vs-ontLvl-1.pdf}

\caption{\label{fig-consist-vs-ontLvl}Relationship between the
consistency of GPT-4 annotations (using the stringent criterion) and the
level of each phenotype within the HPO ontology (with the number of
phenotypes in parentheses). Greater ontology levels (x-axis) indicate
more specific phenotypes. The subtitle indicates summary statistics for
the overall relationship between HPO level and the proportion of
phenotypes that were annotated consistently. The p-values above each bar
indicate whether the distribution of consistent/inconsistent
annotations, within a given HPO level, significantly deviate from the
expected null distribution.}

@@

\phantomsection\label{cell-fig-metric-corplot}
\begin{figure}[H]

\includegraphics{manuscript_files/figure-pdf/fig-metric-corplot-1.pdf}

\caption{\label{fig-metric-corplot}Pearson correlations between each
individual severity metric and the composite severity score
(`severity\_score\_gpt').}

\phantomsection\label{cell-fig-severity-class}
\begin{figure}[H]

\includegraphics{manuscript_files/figure-pdf/fig-severity-class-1.pdf}
